The brightness and spatial distributions of terrestrial radio sources

### 2012-11-22 articles/Article-RFI-distribution/article-rfi-distribution.tex
Finished article

\documentclass[useAMS,usenatbib]{mn2e} 
%\documentclass{article}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[font={it},labelfont={bf}]{caption}
%\usepackage[round]{natbib}
\usepackage{units}
%\usepackage[utopia]{mathdesign}
\usepackage{times} %A
%\usepackage{fourier} %B
%\usepackage{lmodern} % C
%\usepackage{mathpazo} % D
%\usepackage[T1]{fontenc}
%\usepackage{euler} % D
\usepackage{color}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}

@@

\begin{abstract}
Feeble undetected sources of radio-frequency interference (RFI) might become visible after integration of long radio observations if they are coherently present over time. Thereby, they might occlude the astronomical signal of interest. This issue is especially important for epoch of reionization (EoR) projects that try to detect the very faint redshifted HI signals of the earliest objects in the Universe. We explore the low-frequency RFI situation by studying the $\log N$--$\log S$ histograms of data observed with LOFAR. The histograms shows significant features, of which a power law with estimated exponent around -1.6 to -1.5 is most striking. With several assumptions, this can be explained with a uniform distribution of terrestrial radio sources whose radiation follows the Hata propagation model. Extrapolation of the power law would imply that the current LOFAR EoR observations should be severely RFI limited if these sources are coherently present. Since these observations have achieved the thermal (sky) noise limit, a logical implication is that most of the RFI sources do not add coherently.
\end{abstract}

@@

Regular radio observations record one or a few sessions of typical a few hours and results are integrated. In these cases, it often suffices to only excise interference that is apparent and thus above the noise to reach the thermal noise limit of the instrument. A new challenge arises when one desires much deeper observations, and hunderds of hours of observations need to be integrated. In such a case, feeble interference caused by stationary sources might not manifest itself above the noise in individual observations, but is \index{coherency (RFI)} coherently \index{RFI!coherency} present in the data. Subsequently, when averaging these data, the interference might become apparent and occlude the signal of interest. This is very relevant for 21-cm Epoch of Reionisation (EoR) experiments, because these involve long integration times. Several such experiments are underway, both to measure the angular power spectrum \citep{gmrt-eor-2011-paciga,de-bruyn-eor-ursi-2011,eor-paper-2011-jacobs,eor-mwa-2012-williams} or global signal \citep{edges}. For these experiments, it is important to know the possible effect of low-level interferences on the data, as these might overshadow or alter the signal of interest. In this article, we will connect new insights about RFI to the angular Epoch of Reionisation experiment that is using the Low-Frequency Array \citep{de-bruyn-eor-ursi-2011}.

This work explores the information that is present in interference distributions, in order to analyse possible low-level interference that is not detectable by standard detection methods. Our approach is similar to the radio source count analysis that is used in cosmology \citep{vonhoerner-radio-source-counts}, also named \index{log N--log S plot} $\log N$ -- $\log S$ analysis, where $N$ and $S$ refer to the celestial source count and brightness respectively. The slope in such a plot contains information about source populations, their luminosity functions and the geometry of the Universe. We analyse such a double-logarithmic plot for the case of terrestrial sources, with the ultimate goal of predicting their full spatial and brightness distributions. This results in a better insight in the effects of low-level interference and allows one to simulate the effects of interference more accurately. A distribution can be expressed either as a differential or cumulative function. This work uses differential distributions, because they are better suited for visualizing the parameters we are interested in, although the two basically convey the same information.

@@

\section{Predicting the brightness distribution} \label{sec:distribution-prediction}
\index{brightness distribution}\index{spatial distribution}\index{power law}\index{transmitters!distribution}
Interference is generated by many different kind of transmitters, and these will have different spatial and brightness distributions (``spatial'' refers here to the distribution on the Earth). For example, \index{aeroplanes} aeroplanes and \index{satellite} satellites have widely different heights, while other sources are ground-based. Even ground-based sources might be spread differently. For example, it can be expected that \index{CB devices} citizens' band (CB) devices, that are often used in cars, are differently distributed from broadcasting transmitters. The frequencies of deliberately transmitting sources of a particular class of devices are constrained by the bands that have been allocated for the given class, which might allow one to distinguish different sources to some extent.

\index{morphology}
In a time-frequency plot, interfering sources can have complex structures. They can also be intermittent and different sources might overlap in time-frequency space. An example of interfering sources can be seen in Fig.~\ref{fig:dist-rfi-example}, which shows raw visibility data of one baseline of a LOFAR observation in a time-frequency diagram (dynamic spectrum). Because many sources change over time, are repetitive or affect multiple channels, many sources produce multiple unconnected features in the time-frequency diagram. It is often not clear what constitutes a single interfering source, hence it is hard to count individual sources. Instead, we will count the number of times a given brightness occurs in time-frequency space. This --- as well as many other effects --- will of course influence the distribution. First we will derive the expected intrinsic source distribution. After that, we will analyse the issues that arise when measuring the distribution by counting samples.

@@

In every dynamic spectrum we can measure the number of times that the flux density is within a particular range. Dividing this quantity by the total number of samples yields the relative number of events as a function of intensity. We will refer to this quantity with the statistical term \index{frequency density} ``frequency density'' (not to be confused with the physical radiation frequency). We will now start by deriving a prediction of the frequency density function of ground-based interfering sources. Consider an interfering point source of strength $I$. This source is observed by an interferometer that consists of two antennae or stations with gains $g_1,g_2$, which include all instrumental effects. The antennae are located at distances $r_1,r_2$ from the source. The interferometer will record an apparent instantaneous strength $\mathcal{S}$ of
\begin{equation}
 \mathcal{S}(r_1, r_2) = I \frac{g_1 g_2}{r_1 r_2},
\end{equation}
with $g_1,g_2>0$ and $r_\textrm{L} > r_1,r_2 > 0$. Here, $r_\textrm{L}$ is a limiting distance, which will be well below the diameter of the Earth. We will limit our analysis to cross-correlated antennae; the auto-correlations will be ignored.

We assume that the source is observed fully coherent, but a possible decoherence factor can be absorbed in the gains. Due to the small bandwidth of most interfering sources, most RFI will be received coherently, because of the \index{narrow-band condition} narrow-band condition. For example, if the bandwidth of the signal $\Delta\nu=1$ kHz, the narrow-band condition $\Delta \nu \ll (2\pi \tau)^{-1}$ with correlation delay $\tau$ will hold for baselines up to a few km, because it holds as long as the baseline length is significantly less than $\Delta x = c / (2 \pi \Delta \nu) \approx$ 50 km.

@@

Therefore, if we plot the histogram of the RFI amplitudes in a $\log$-$\log$ plot, we predict to see a \index{slope (log-log plot)} slope of $-2$ over the interval in which the RFI sources are spread like uniform sources on a two-dimensional plane.

\subsection{Spherical case} \index{spherical distribution}
If we do take into account the fact that the sources lie on a sphere, Eq.~\eqref{eq:cumulative-distance-function} changes. In this case, the number of sources corresponds with the surface area of a hemisphere. By using the formula for the surface area of a hemisphere and some basic geometry, one finds that
\begin{equation}\label{eq:cumulative-spherical}
\tilde{F}_{\textrm{distance} \le r} = c 2\pi R^2\left(1-\cos(2\arcsin \frac{r}{2R})\right) \hspace{0.4cm}\textrm{with } r \le 2R,
\end{equation}
with $R$ the radius of the Earth. Eqs.~\eqref{eq:cumulative-distance-function} and \eqref{eq:cumulative-spherical} have the same value over the range they can be evaluated. Therefore, the spherical and 2-dimensional cases lead to a distribution with equal power law. Evaluating Eq.~\eqref{eq:cumulative-spherical} is limited to $r \le 2R$, but in practice the lowest amplitude of observed sources is further constrained because of the curvature of the Earth. This causes distant sources to settle below the horizon. Nevertheless, some of those sources might still be visible due to reflection of the ionosphere and other (semi-)spherical propagation effects.

\subsection{Propagation effects} \index{propagation (EM)}
So far, we have assumed that the electromagnetic radiation propagates through \index{free space} free space, resulting in a $r^{-2}$ fall-off. In reality, the radiation will be affected by complicated propagational effects. Because of the irregular surface of the Earth (including urban areas) and the absence of a direct \index{line of sight} line of sight between transmitter and receiver, the propagation mode might be indirect. Multiple indirect paths might be formed by reflection, refraction or diffraction of the electromagnetic wave.

A commonly used propagation model is the empirical model determined by \citet{okumura-propagation-model}, which was further developed by \citet{hata-propagation-loss}. \citeauthor{hata-propagation-loss} gives the following analytical estimate for $L_p$, the electromagnetic propagation loss over land:
\begin{align} 
\notag                L_p & = & 69.55 + 26.16 \log_{10} f_c - 13.82 \log_{10} h_b - \\
\label{eq:hata-model} & &  a(h_m) + (44.9 - 6.55 \log_{10} h_b) \log_{10} r, 
\end{align}
where $L_p$ the loss in dB; $f_c$ the radiation frequency in MHz; $h_b$ the height of the transmitting antenna in m; $h_m$ the height of the receiving antenna in m; $r$ the distance between the antennae in m; and $a(h_m)$ a correction factor in dB that corrects for the height of the receiving antenna and the \index{urban density|see{population density}} urban \index{population density} density. \citeauthor{hata-propagation-loss} found this model to be representative for frequencies $f_c \sim$~150--1500~MHz, with \index{transmitters!height} transmitter heights $h_b \sim$~30--200~m, receiver heights $h_m \sim$~1--10~m and over distances $r\sim$~1--20~km.

Collecting the terms of Eq.~\eqref{eq:hata-model} that are not depending on $r$ in a single variable $\zeta$, and converting from a subtrahend in decibels to a flux density factor $L_S$, results in
\begin{equation} \label{eq:propagation-loss-factor}
 L_S = \zeta r^{4.49 - 0.655 \log_{10} h_b},
\end{equation}
with
\begin{equation}
 \zeta = \frac{f_c^{2.616}}{h_b^{1.382}} - 10^{6.955-\frac{1}{10} a(h_m)}.
\end{equation}
The model is based on urban areas, but corrections are given by \citeauthor{hata-propagation-loss} for sub-urban areas with a lower population density and for open areas. These corrections are independent of $r$, thus would only change $\zeta$. Note that according to Hata's model, the exponent of the power law in Eq.~\eqref{eq:propagation-loss-factor} depends only on the height of the transmitting antenna, i.e., it is independent of frequency, receiver height and urban density. Now, if in Eqs.~\eqref{eq:cumulative-rfi-distribution} and \eqref{eq:two-dimensional-distribution} one replaces the definition of $S(r)$ from Eq.~\eqref{eq:amplitude-fall-off-in-free-space} with one that includes the propagation effects,
\begin{equation}
 S(r) = \frac{\zeta Ig}{r^\eta},
\end{equation}
with $\eta = 0.655 \log_{10} h_b - 4.49$, one finds the frequency density function $f_p$ that considers propagation effects,
% S(r):=I*g/(r^alpha);
% assume(g>0,I>0,S>0);
% diff(c*%pi*r^2,S),solve(S(r)=S,r);
\begin{align}
f_p(S)
= & \frac{d}{dS} c \pi \left( \frac{\zeta Ig }{S} \right)^{2/\eta}
= \frac{c 2\pi}{\eta S} \left( \frac{\zeta Ig}{S} \right)^{2/\eta}.
\end{align}
Consequently, due to non-free space propagation effects, the observed log-log histogram is predicted to have a $\frac{2}{\eta}-1$ slope. By substituting $\eta$, one finds
\begin{equation}
\textrm{slope}(h_b) = \frac{1}{0.3275 \log_{10} h_b - 2.245} - 1.
\end{equation}
\begin{figure}
\begin{center}\hspace{-1cm}\includegraphics[width=6cm]{img/transmitter-height/plot-transmitter-height-trimmed}
\caption{Effect of transmitter height on the slope of a log-log histogram.}
\label{fig:plot-transmitter-height}
\end{center}
\end{figure}
This is valid for \index{transmitters!height} transmitters that have a height of 30--200 m, the range over which Hata's model was defined. This yields estimated distribution slopes of $-1.57$ and $-1.67$ for $30$ m and $200$ m high transmitters respectively. In Figure~\ref{fig:plot-transmitter-height}, the slope value is plotted as a function of the transmitter height, including extrapolated values for transmitter heights down to 1 m.

\subsection{Noise contribution} \label{sec:histogram-noise}\index{noise}
The full measured distribution will consist of the power-law distribution combined with that of the noise and the celestial signal. For now, we will ignore the contribution of the celestial signal, as its contribution to the amplitude distribution will be minimal when observing fields without strong sources. Noise, however, will have a contribution. The real and imaginary components of receiver noise are independent and identically Gaussian distributed. Consequently, an amplitude $x$ will be Rayleigh distributed: \index{Rayleigh distribution}
\begin{equation}\label{eq:rayleigh-formula}
f_\textrm{noise}(x) =
\begin{cases}
\frac{x}{\sigma^2} e^{\frac{-x^2}{2\sigma^2}} & x > 0, \\
0 & \textrm{otherwise.}
\end{cases}
\end{equation}
Because most of the samples will be unaffected by RFI, this will be the dominating distribution. The Rayleigh distribution is plotted together with the -2 power-law distribution of Eq.~\eqref{eq:two-dimensional-distribution} in Fig.~\ref{fig:rayleigh-and-rfi-distributions}.

@@

To understand the RFI-with-noise distribution, one needs to consider the individual contributions of the real and imaginary components. The real and imaginary components of samples that are affected by RFI, will follow a \index{bivariate distribution} bivariate distribution: each complex component is the sum of a complex component of RFI and a noise sample. As said, the noise comes from a Gaussian distribution with zero mean. The RFI distribution is assumed to be a power law with yet unknown exponent.

To calculate the corresponding amplitude distribution that is formed by the combination of the real and imaginary component distributions, it is easier to perform numerical simulations, as algebraic calculation is not trivial. The histogram can be numerically estimated by drawing complex real and imaginary samples from the two distributions and calculating and counting the amplitudes. A sample can be drawn from the RFI distribution by normalizing and inverting the cumulative frequency function in Eq.~\eqref{eq:cumulative-rfi-distribution} and evaluating it for a uniformly distributed variable. Note however that $F_{\textrm{amplitude}\le S}(S)$ is not limited; when decreasing the amplitude $S$ towards zero, the number of sources will go to infinity. Therefore, to \index{drawing samples} draw a sample by using the \index{inversion method} inversion method, this would require a uniformly distributed variable $\sim U(0,\infty)$. By assuming that there are no samples beyond some limiting distance $r_L$, then $F_{\textrm{amplitude}\le S}(S) = N-c \pi r_L^2$ for $S < Ig/r_L^2$. Consequently, by sampling a uniform value $x_u$ with $0 \le x_u \le N-c \pi r_L^2$ and applying the inverse of Eq.~\eqref{eq:cumulative-rfi-distribution}, we will sample amplitudes $S$ with $S \ge Ig/r_L^2$, and $S\sim$ the power-law distribution.

A single complex RFI contaminated sample $S_{\textrm{RFI}}$ can then be sampled with
\begin{equation} \label{eq:drawing-rfi-distribution}
\textrm{real}(S_{\textrm{RFI}})\leftarrow \frac{Ig}{x_u r_L^2}\textrm{,\hspace{1mm} imag}(S_{\textrm{RFI}}) \leftarrow \frac{Ig}{x_u r_L^2},
\end{equation}
with $x_u \sim U(0, 1)$, and the corresponding formula for drawing a sample $S$ that is contaminated by both RFI and noise is
\begin{equation} \label{eq:drawing-combined-distribution}
\textrm{real}(S) \leftarrow x_n + \frac{Ig}{x_u r_L^2}
\textrm{,\hspace{1mm} imag}(S) \leftarrow x_n + \frac{Ig}{x_u r_L^2},
\end{equation}
with $x_n \sim N(\mu=0; \sigma=1)$. Of course, the final amplitude sample $S_A$ can be calculated with 
\begin{equation} \label{eq:drawing-amplitudes}
S_A \leftarrow |S| = \sqrt{\textrm{real}({S})^2+\textrm{imag}({S})^2}.
\end{equation}
An example of distribution curves for various settings of $Ig/r_L^2$ is given in Fig.~\ref{fig:rayleigh-and-rfi-combined}.

@@

One instrumental effect that is absorbed in $g$ is the \index{frequency response|see{band-pass}} frequency response of the instrument, i.e., the antenna response in combination with the passband of the analogue and digital filters. Because the data that are analysed in Sect.~\ref{sec:dist-results} have initially not been band-pass calibrated, the instrumental response is not uniform over frequency. We determined that the variation due to the \index{band-pass} band-pass is about one order of magnitude in the LBA and about a factor of two in the HBA.

@@

\index{beam}\index{LOFAR!beam}
Another effect that is absorbed in $g$, is the beam of the instrument. At the point of writing, LOFAR beam models are still being developed and are not yet well parametrized. However, since it is likely that most RFI sources are observed at the edges of the beam, it can be expected that the beam will have a benign effect on the histogram of an observation, comparable with the effect of the frequency response.

@@

\subsection{Creating a histogram} \label{sec:histogram}\index{histogram}
While creating a histogram is trivial, it is important to note that it is necessary to have a variable bin size. This is mandated by the large dynamic range of the histogram that we are interested in. Therefore, we chose to have a bin size that increases linearly with the amplitude $S$, and the frequency distribution is normalized by the bin size after counting. Consequently, in parts of the histogram that have a sparse number of samples, the outlying samples will follow a $1/S$ curve, or a -1 slope in a log-log plot. This can be seen in the tails of the distributions of Fig.~\ref{fig:rayleigh-and-rfi-combined}. This, however, is not an intrinsic feature of the data but a consequence of the \index{binning} binning method.

>>>>

\documentclass[useAMS,usenatbib]{mn2e} 
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[font={it},labelfont={bf}]{caption}
\usepackage{units}
\usepackage{times}
\usepackage{color}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}

@@

\begin{abstract}
Feeble undetected sources of radio-frequency interference (RFI) might become visible after integration of long radio observations if they are coherently present over time. Thereby, they might occlude the astronomical signal of interest. This issue is especially important for epoch of reionization (EoR) projects that try to detect the faint redshifted HI signals of the earliest objects in the Universe. We explore the low-frequency RFI situation by studying the $\log N$--$\log S$ histograms of data observed with LOFAR. The histograms shows a RFI distribution that follows a power-law distribution with an estimated exponent around -1.6 to -1.5. With several assumptions, this can be explained with a uniform distribution of terrestrial radio sources whose radiation follows existing propagation models. Extrapolation of the power law would imply that the current LOFAR EoR observations should be severely RFI limited if these sources are coherently present. Since these observations have achieved the thermal (sky) noise limit, a logical implication is that most of the RFI sources do not add coherently.
\end{abstract}

@@

Regular radio observations record one or a few sessions of typical a few hours and results are integrated. In these cases, it often suffices to only excise interference that is apparent and thus above the noise to reach the thermal noise limit of the instrument. A new challenge arises when one desires much deeper observations, and hundreds of hours of observations need to be integrated. In such a case, feeble interference caused by stationary sources might not manifest itself above the noise in individual observations, but is coherently present in the data. Subsequently, when averaging these data, the interference might become apparent and occlude the signal of interest. This is very relevant for 21-cm Epoch of Reionisation (EoR) experiments, because these involve long integration times. Several such experiments are underway, both to measure the angular power spectrum \citep{gmrt-eor-2011-paciga,de-bruyn-eor-ursi-2011,eor-paper-2011-jacobs,eor-mwa-2012-williams} or global signal \citep{edges}. For these experiments, it is important to know the possible effect of low-level interferences on the data, as these might overshadow or alter the signal of interest. In this article, we will connect new insights about RFI to the angular Epoch of Reionisation experiment that is using the Low-Frequency Array \citep{de-bruyn-eor-ursi-2011}.

This work explores the information that is present in interference distributions, in order to analyse possible low-level interference that is not detectable by standard detection methods. Our approach is similar to the radio source count analysis that is used in cosmology \citep{vonhoerner-radio-source-counts}, also named $\log N$ -- $\log S$ analysis, where $N$ and $S$ refer to the celestial source count and brightness respectively. The slope in such a plot contains information about source populations, their luminosity functions and the geometry of the Universe. We analyse such a double-logarithmic plot for the case of terrestrial sources, with the ultimate goal of predicting their full spatial and brightness distributions. This results in a better insight in the effects of low-level interference and allows one to simulate the effects of interference more accurately. A distribution can be expressed either as a differential or cumulative function. This work uses differential distributions, because they are better suited for visualizing the parameters we are interested in, although the two basically convey the same information.

@@

\section{Predicting the brightness distribution} \label{sec:distribution-prediction}
Interference is generated by many different kind of transmitters, and these will have different spatial and brightness distributions (``spatial'' refers here to the distribution on the Earth). For example, aeroplanes and satellites have widely different heights, while other sources are ground-based. Even ground-based sources might be spread differently. For example, it can be expected that citizens' band (CB) devices, that are often used in cars, are differently distributed from broadcasting transmitters. The frequencies of deliberately transmitting sources of a particular class of devices are constrained by the bands that have been allocated for the given class, which might allow one to distinguish different sources to some extent.

In a time-frequency plot, interfering sources can have complex structures. They can also be intermittent and different sources might overlap in time-frequency space. An example of interfering sources can be seen in Fig.~\ref{fig:dist-rfi-example}, which shows raw visibility data of one baseline of a LOFAR observation in a time-frequency diagram (dynamic spectrum). Because many sources change over time, are repetitive or affect multiple channels, many sources produce multiple unconnected features in the time-frequency diagram. It is often not clear what constitutes a single interfering source, hence it is hard to count individual sources. Instead, we will count the number of times a given brightness occurs in time-frequency space. This --- as well as many other effects --- will of course influence the distribution. First we will derive the expected intrinsic source distribution. After that, we will analyse the issues that arise when measuring the distribution by counting samples.

@@

In every dynamic spectrum we can measure the number of times that the flux density is within a particular range. Dividing this quantity by the total number of samples yields the relative number of events as a function of intensity. We will refer to this quantity with the statistical term ``frequency density'' (not to be confused with the physical radiation frequency). We will now start by deriving a prediction of the frequency density function of ground-based interfering sources. Consider an interfering point source of strength $I$. This source is observed by an interferometer that consists of two antennae or stations with gains $g_1,g_2$, which include all instrumental effects. The antennae are located at distances $r_1,r_2$ from the source. The interferometer will record an apparent instantaneous strength $\mathcal{S}$ of
\begin{equation}
 \mathcal{S}(r_1, r_2) = I \frac{g_1 g_2}{r_1 r_2},
\end{equation}
with $g_1,g_2>0$ and $r_\textrm{L} > r_1,r_2 > 0$. Here, $r_\textrm{L}$ is a limiting distance, which will be well below the diameter of the Earth. We will limit our analysis to cross-correlated antennae; the auto-correlations will be ignored.

We assume that the source is observed fully coherent, but a possible decoherence factor can be absorbed in the gains. Due to the small bandwidth of most interfering sources, most RFI will be received coherently, because of the narrow-band condition. For example, if the bandwidth of the signal $\Delta\nu=1$ kHz, the narrow-band condition $\Delta \nu \ll (2\pi \tau)^{-1}$ with correlation delay $\tau$ will hold for baselines up to a few km, because it holds as long as the baseline length is significantly less than $\Delta x = c / (2 \pi \Delta \nu) \approx$ 50 km.

@@

Therefore, if we plot the histogram of the RFI amplitudes in a $\log$-$\log$ plot, we predict to see a slope of $-2$ over the interval in which the RFI sources are spread like uniform sources on a two-dimensional plane. Therefore, RFI follows a power-law distribution.

\subsection{Spherical case}
If we do take into account the fact that the sources lie on a sphere instead of a plane, Eq.~\eqref{eq:cumulative-distance-function} changes. In this case, the number of sources corresponds with the surface area of a hemisphere. By using the formula for the surface area of a hemisphere and some basic geometry, one finds that
\begin{equation}\label{eq:cumulative-spherical}
\tilde{F}_{\textrm{distance} \le r} = c 2\pi R^2\left(1-\cos(2\arcsin \frac{r}{2R})\right) \hspace{0.4cm}\textrm{with } r \le 2R,
\end{equation}
with $R$ the radius of the Earth. Eqs.~\eqref{eq:cumulative-distance-function} and \eqref{eq:cumulative-spherical} have the same value over the range they can be evaluated. Therefore, the spherical and 2-dimensional cases lead to a distribution with equal power law. Evaluating Eq.~\eqref{eq:cumulative-spherical} is limited to $r \le 2R$, but in practice the lowest amplitude of observed sources is further constrained because of the curvature of the Earth. This causes distant sources to settle below the horizon. Nevertheless, some of those sources might still be visible due to reflection of the ionosphere and other (semi-)spherical propagation effects.

\subsection{Propagation effects}
So far, we have assumed that the electromagnetic radiation propagates through free space, resulting in a $r^{-2}$ fall-off. In reality, the radiation will be affected by complicated propagational effects. Because of the irregular surface of the Earth (including urban areas) and the absence of a direct line of sight between transmitter and receiver, the propagation mode might be indirect. Multiple indirect paths might be formed by reflection, refraction or diffraction of the electromagnetic wave.

A commonly used propagation model is the empirical model determined by \citet{okumura-propagation-model}, which was further developed by \citet{hata-propagation-loss}. \citeauthor{hata-propagation-loss} gives the following analytical estimate for $L_p$, the electromagnetic propagation loss over land:
\begin{align} 
\notag                L_p & = & 69.55 + 26.16 \log_{10} f_c - 13.82 \log_{10} h_b - \\
\label{eq:hata-model} & &  a(h_m) + (44.9 - 6.55 \log_{10} h_b) \log_{10} r, 
\end{align}
where $L_p$ the loss in dB; $f_c$ the radiation frequency in MHz; $h_b$ the height of the transmitting antenna in m; $h_m$ the height of the receiving antenna in m; $r$ the distance between the antennae in m; and $a(h_m)$ a correction factor in dB that corrects for the height of the receiving antenna and the urban density. \citeauthor{hata-propagation-loss} found this model to be representative for frequencies $f_c \sim$~150--1500~MHz, with transmitter heights $h_b \sim$~30--200~m, receiver heights $h_m \sim$~1--10~m and over distances $r\sim$~1--20~km.

Converting from a subtrahend in decibels to a flux density factor $L_S$, and collecting the terms of Eq.~\eqref{eq:hata-model} that are not depending on $r$ in a single variable $\zeta$, results in
\begin{equation} \label{eq:propagation-loss-factor}
 L_S = \zeta r^{4.49 - 0.655 \log_{10} h_b},
\end{equation}
with
\begin{equation}
 \zeta = \frac{f_c^{2.616}}{h_b^{1.382}} - 10^{6.955-\frac{1}{10} a(h_m)}.
\end{equation}
The model is based on urban areas, but corrections are given by \citeauthor{hata-propagation-loss} for sub-urban areas with a lower population density and for open areas. These corrections are independent of $r$, thus would only change $\zeta$. Note that according to Hata's model, the exponent of the power law in Eq.~\eqref{eq:propagation-loss-factor} depends only on the height of the transmitting antenna, i.e., it is independent of frequency, receiver height and urban density. Now, if in Eqs.~\eqref{eq:cumulative-rfi-distribution} and \eqref{eq:two-dimensional-distribution} one replaces the definition of $S(r)$ from Eq.~\eqref{eq:amplitude-fall-off-in-free-space} with one that includes the propagation effects,
\begin{equation}
 S(r) = \frac{\zeta Ig}{r^\eta},
\end{equation}
with $\eta = 0.655 \log_{10} h_b - 4.49$, one finds the frequency density function $f_p$ that considers propagation effects,
% S(r):=I*g/(r^alpha);
% assume(g>0,I>0,S>0);
% diff(c*%pi*r^2,S),solve(S(r)=S,r);
\begin{align}
f_p(S)
= & \frac{d}{dS} c \pi \left( \frac{\zeta Ig }{S} \right)^{2/\eta}
= \frac{c 2\pi}{\eta S} \left( \frac{\zeta Ig}{S} \right)^{2/\eta}.
\end{align}
Consequently, due to non-free space propagation effects, the observed log-log histogram is predicted to have a $\frac{2}{\eta}-1$ slope. By substituting $\eta$, one finds
\begin{equation}
\textrm{slope}(h_b) = \frac{1}{0.3275 \log_{10} h_b - 2.245} - 1.
\end{equation}
\begin{figure}
\begin{center}\hspace{-1cm}\includegraphics[width=6cm]{img/transmitter-height/plot-transmitter-height-trimmed}
\caption{Effect of transmitter height on the slope of a log-log histogram.}
\label{fig:plot-transmitter-height}
\end{center}
\end{figure}
This is valid for transmitters that have a height of 30--200 m, the range over which Hata's model was defined. This yields estimated distribution slopes of $-1.57$ and $-1.67$ for $30$ m and $200$ m high transmitters respectively. In Figure~\ref{fig:plot-transmitter-height}, the slope value is plotted as a function of the transmitter height, including extrapolated values for transmitter heights down to 1 m.

\subsection{Noise contribution} \label{sec:histogram-noise}
The full measured distribution will consist of the power-law distribution combined with that of the noise and the celestial signal. For now, we will ignore the contribution of the celestial signal, as its contribution to the amplitude distribution will be minimal when observing fields without strong sources. Noise, however, will have a contribution. The real and imaginary components of receiver noise are independent and identically Gaussian distributed. Consequently, an amplitude $x$ will be Rayleigh distributed: 
\begin{equation}\label{eq:rayleigh-formula}
f_\textrm{noise}(x) =
\begin{cases}
\frac{x}{\sigma^2} e^{\frac{-x^2}{2\sigma^2}} & x > 0, \\
0 & \textrm{otherwise.}
\end{cases}
\end{equation}
Because most of the samples will be unaffected by RFI, this will be the dominating distribution. The Rayleigh distribution is plotted together with the -2 power-law distribution of Eq.~\eqref{eq:two-dimensional-distribution} in Fig.~\ref{fig:rayleigh-and-rfi-distributions}.

@@

To understand the distribution of samples consisting of RFI + noise, one needs to consider the individual contributions of the real and imaginary components. The real and imaginary components of samples that are affected by RFI, will follow a bivariate distribution: each complex component is the sum of a complex component of RFI and a noise sample. The noise comes from a Gaussian distribution with zero mean. The RFI distribution is assumed to be a power-law distribution with yet unknown exponent.

To calculate the corresponding amplitude distribution that is formed by the combination of the real and imaginary component distributions, it is easier to perform numerical simulations, as algebraic calculation is not trivial. The histogram can be numerically estimated by drawing complex real and imaginary samples from the two distributions and calculating and counting the amplitudes. A sample can be drawn from the RFI distribution by normalizing and inverting the cumulative frequency function in Eq.~\eqref{eq:cumulative-rfi-distribution} and evaluating it for a uniformly distributed variable. Note however that $F_{\textrm{amplitude}\le S}(S)$ is not limited; when decreasing the amplitude $S$ towards zero, the number of sources will go to infinity. Therefore, to draw a sample by using the inversion method, this would require a uniformly distributed variable $\sim U(0,\infty)$. By assuming that there are no samples beyond some limiting distance $r_L$, then $F_{\textrm{amplitude}\le S}(S) = N-c \pi r_L^2$ for $S < Ig/r_L^2$. Consequently, by sampling $x_u$ from a uniform distribution with $0 \le x_u \le N-c \pi r_L^2$ and applying the inverse of Eq.~\eqref{eq:cumulative-rfi-distribution}, we will sample amplitudes $S$ with $S \ge Ig/r_L^2$, and $S\sim$ the power-law distribution.

By normalizing $x_u$, a single complex RFI contaminated sample $S_{\textrm{RFI}}$ can be sampled with
\begin{equation} \label{eq:drawing-rfi-distribution}
\textrm{real}(S_{\textrm{RFI}})\leftarrow \frac{Ig}{x_u r_L^2}\textrm{,\hspace{1mm} imag}(S_{\textrm{RFI}}) \leftarrow \frac{Ig}{y_u r_L^2},
\end{equation}
with $x_u, y_u \sim U(0, 1)$, and the corresponding formula for drawing a sample $S$ that is contaminated by both RFI and noise is
\begin{equation} \label{eq:drawing-combined-distribution}
\textrm{real}(S) \leftarrow x_n + \frac{Ig}{x_u r_L^2}
\textrm{,\hspace{1mm} imag}(S) \leftarrow y_n + \frac{Ig}{y_u r_L^2},
\end{equation}
with $x_n \sim N(\mu=0; \sigma=1)$. Of course, the final amplitude sample $S_A$ can be calculated with 
\begin{equation} \label{eq:drawing-amplitudes}
S_A \leftarrow |S| = \sqrt{\textrm{real}({S})^2+\textrm{imag}({S})^2}.
\end{equation}
An example of distribution curves for various settings of $Ig/r_L^2$ is given in Fig.~\ref{fig:rayleigh-and-rfi-combined}.

@@

One instrumental effect that is absorbed in $g$ is the frequency response of the instrument, i.e., the antenna response in combination with the passband of the analogue and digital filters. Because the data that are analysed in Sect.~\ref{sec:dist-results} have initially not been band-pass calibrated, the instrumental response is not uniform over frequency. We determined that the variation due to the band-pass is about one order of magnitude in the LBA and about a factor of two in the HBA.

@@

Another effect that is absorbed in $g$, is the beam of the instrument. At the point of writing, LOFAR beam models are still being developed and are not yet well parametrized. However, since it is likely that most RFI sources are observed at the edges of the beam, it can be expected that the beam will have a benign effect on the histogram of an observation, comparable with the effect of the frequency response.

@@

\subsection{Creating a histogram} \label{sec:histogram}
While creating a histogram is trivial, it is important to note that it is necessary to have a variable bin size. This is mandated by the large dynamic range of the histogram that we are interested in. Therefore, we chose to have a bin size that increases linearly with the amplitude $S$, and the frequency distribution is normalized by the bin size after counting. Consequently, in parts of the histogram that have a sparse number of samples, the outlying samples will follow a $1/S$ curve, or a -1 slope in a log-log plot. This can be seen in the tails of the distributions of Fig.~\ref{fig:rayleigh-and-rfi-combined}. This, however, is not an intrinsic feature of the data but a consequence of the binning method.

\subsection{Estimating $\sigma$ and slope parameters}
The mode $\hat\sigma$ of the Rayleigh distribution is estimated by finding the amplitude with the maximum occurrences, i.e., the amplitude corresponding to the peak of the histogram. The slope is estimated using linear regression over a visually selected interval.
%\begin{equation}
%R=\left[20\hat\sigma; 10^{\frac{1}{2}\left(\log_{10} 20\hat\sigma + \log_{10} S_\textrm{max}\right) } \right].
%\end{equation}
%In terms of the log-log histogram, the start of the interval corresponds with the point that is $1.3$ units to the right of the peak in the histogram, and the interval end corresponds with the point halfway between the start point and the maximum amplitude. This interval starts near the start of the RFI slope but past where the Rayleigh curve dominates. It also does not include the noisy data at the end of the curve. The curves are not always completely straight in this interval, but since we use the same stable way to determine the interval over which to calculate the slope, we can at least compare them.
